% ============================================================
%  AgriDrone — Springer LNCS format
%  Target: CVPPA 2026 @ ECCV / PlantCLEF 2026 @ CLEF
%  Compile: pdflatex main && bibtex main && pdflatex main && pdflatex main
% ============================================================
\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{multirow}
\usepackage{array}
\renewcommand\UrlFont{\color{blue}\rmfamily}

% ── FILL IN after Zenodo upload ──────────────────────────────
\newcommand{\codeurl}{https://github.com/Ashut0sh-mishra/agri-drone}
\newcommand{\zenododoi}{TODO: 10.5281/zenodo.XXXXXXX}   % <-- replace after upload
% ─────────────────────────────────────────────────────────────

\begin{document}

\title{When Do Hand-Authored Symbolic Rules Help Modern CNNs\\
       for Agricultural Disease Classification?\\
       A Negative Result}

\titlerunning{Symbolic Rules for Crop-Disease CNNs: A Negative Result}

\author{Ashutosh Mishra}

\authorrunning{A. Mishra}

\institute{Independent Researcher, India\\
\email{mishra.ashutosh@gmail.com}\\
\url{\codeurl}}

\maketitle

% ── Abstract ─────────────────────────────────────────────────
\begin{abstract}
Hybrid neuro-symbolic systems---a CNN classifier audited by a hand-authored
symbolic rule engine---are a recurring proposal in precision-agriculture
literature, typically justified by appeals to interpretability and farmer
trust. We test this proposal on 934 stratified held-out smartphone images
covering 21 rice- and wheat-disease classes and report a \textbf{negative
result}. A standalone YOLOv8n-cls backbone (Config~A) achieves
96.15\,\% accuracy, macro-F1\,=\,0.9618, MCC\,=\,0.9596
(95\,\% bootstrap CI [0.9486, 0.9732]) at 15.4\,ms/image on a
single T4~GPU. Adding an agronomist-informed rule layer (Config~B) changes
only \textbf{4 of 934 predictions}---all four in the wrong direction---
yielding McNemar $p = 0.134$ (not significant), a 29$\times$ latency tax
(444\,ms), and an 840\,\% increase in expected monetary loss under a cost
matrix calibrated from ICAR/DAE advisory tables. Per-class McNemar tests
with Holm--Bonferroni correction confirm 0/21 classes show a significant
A-vs-B difference. A rules-only ablation (Config~C) collapses to 13.4\,\%
accuracy, confirming the rule engine is non-trivial: the issue is
\emph{informational redundancy} with the CNN. For clean-to-moderately-noisy
smallholder imagery at current CNN accuracy ceilings, hand-authored symbolic
rules add cost without adding recognition or cost-adjusted safety.

\keywords{Negative result \and Crop-disease classification \and
Neuro-symbolic ablation \and YOLOv8 \and Expected monetary loss \and
Smallholder agriculture}
\end{abstract}

% ── 1. Introduction ──────────────────────────────────────────
\section{Introduction}

Precision agriculture increasingly deploys deep-learning classifiers for
automated crop-disease triage from smartphone or drone imagery. A persistent
theme in the literature is the proposal to augment CNN predictions with
hand-authored symbolic rules---typically encoding agronomic knowledge about
disease symptoms (colour signatures, lesion morphology, spatial
patterns)---to improve accuracy, interpretability, or farmer trust
\cite{Ferentinos2018,Ramcharan2017}.

The implicit assumption is that symbolic domain knowledge adds value on top
of what the CNN learns from data alone. We test this assumption directly with
a three-configuration ablation:

\begin{itemize}
  \item \textbf{Config~A:} YOLOv8n-cls only (softmax argmax)
  \item \textbf{Config~B:} YOLOv8n-cls + hand-authored rule engine (weighted re-ranking)
  \item \textbf{Config~C:} Rule engine only (no CNN)
\end{itemize}

Our headline finding is negative. On a 934-image, 21-class held-out test set:
(1) Config~A achieves 96.15\,\% accuracy (95\,\% CI [0.9486, 0.9732]);
(2) Config~B changes only 4 of 934 predictions, all adversely (McNemar
$p = 0.134$);
(3) Config~B inflates expected monetary loss by 840\,\%;
(4) Config~C achieves 13.4\,\%---genuine but informationally redundant signal.

\paragraph{Contributions.}
\begin{enumerate}
  \item A three-configuration ablation with paired statistical tests showing
        rules do not help a strong CNN on this task.
  \item An EML cost analysis showing the hybrid pattern actively harms
        cost-adjusted outcomes.
  \item An open-source rule engine that serves as a non-trivial baseline for
        future neuro-symbolic comparisons.
  \item Reproducible evaluation scripts and committed prediction artefacts
        for independent verification (DOI: \zenododoi).
\end{enumerate}

% ── 2. Related Work ──────────────────────────────────────────
\section{Related Work}

Modern crop-disease classification has advanced rapidly. Mohanty et
al.~\cite{Mohanty2016} demonstrated $>$99\,\% accuracy on PlantVillage lab
imagery, but field performance typically degrades to $\sim$30\,\%
\cite{Barbedo2018}. Hybrid approaches combining CNNs with domain-specific
rules have been proposed repeatedly \cite{Ferentinos2018,Ramcharan2017}, but
head-to-head ablations quantifying the marginal value of the rule layer are
rare. The key gap is empirical: most neuro-symbolic crop-disease papers
report only the hybrid's performance, not a paired comparison against the CNN
alone with statistical testing. We address this gap directly.

% ── 3. Data ──────────────────────────────────────────────────
\section{Data}

\paragraph{Training corpus.}
Training images were sourced from publicly available datasets including
PlantVillage, Kaggle Bangladesh-rice-leaf and Indian-wheat-leaf collections,
covering 21 classes (2 healthy + 5 rice disease + 14 wheat disease/pest).

\paragraph{Held-out evaluation set.}
The test set comprises \textbf{934 images} across 21 classes, split
deterministically at seed~42 with a 70/15/15 train/val/test ratio. Class
sizes range from 20 to 45 images per class.

\paragraph{Augmentation leakage check.}
21.2\,\% of test images carry \texttt{aug\_*} filenames. Every augmented
file has a unique base ID; no sibling augmentations cross split boundaries
(verified in \texttt{LEAKAGE\_INVESTIGATION\_REPORT.md}, committed to the
repository).

% ── 4. Method ────────────────────────────────────────────────
\section{Method}

\subsection{Three-Configuration Ablation}

\begin{table}[h]
\centering
\caption{Ablation configurations.}
\label{tab:configs}
\begin{tabular}{@{}llll@{}}
\toprule
Config & CNN & Rules & Decision \\
\midrule
A & YOLOv8n-cls (1.4M params) & Off  & Softmax argmax \\
B & YOLOv8n-cls               & On   & Weighted re-ranking \\
C & Off                        & On   & Rule scores only \\
\bottomrule
\end{tabular}
\end{table}

\noindent All three configs are evaluated on the same images in the same order.

\subsection{Rule Engine}

The rule engine (\texttt{src/agridrone/vision/rule\_engine.py}) is a
substantive multi-stage system:
\begin{itemize}
  \item \textbf{Colour rules:} HSV-signature matching against per-disease profiles.
  \item \textbf{Texture rules:} Bleaching-ratio detection, spot/pustule scoring.
  \item \textbf{Spatial rules:} Stripe-vs-spot penalisation using Hough-line directionality.
  \item \textbf{Conflict resolution:} \texttt{CandidateScore} aggregation with
        \texttt{ConflictReport} when CNN top-1 $\neq$ rule top-1.
\end{itemize}

Blending weights: $\alpha_{\text{cnn}}=0.70$, $\alpha_{\text{rules}}=0.30$
with evidence; 0.85/0.15 without.

\subsection{Expected Monetary Loss (EML)}

Per-image EML uses cost parameters from ICAR and DAE advisory tables:
$c_{\text{miss}}=\text{\rupee}12{,}000$/ha for critical diseases;
$c_{\text{alarm}}=\text{\rupee}640$/ha (one wasted spray round).

\subsection{Statistical Testing}
\begin{itemize}
  \item \textbf{Bootstrap:} 10{,}000-replicate percentile CIs for accuracy,
        macro-F1, MCC per config.
  \item \textbf{McNemar:} Continuity-corrected paired test ($\alpha=0.05$)
        on image-level correct/incorrect vectors.
  \item \textbf{Holm--Bonferroni:} Per-class McNemar across 21 classes.
\end{itemize}

% ── 5. Results ───────────────────────────────────────────────
\section{Results}

\subsection{Headline Metrics}

\begin{table}[h]
\centering
\caption{Per-configuration performance on 933 successfully predicted images.}
\label{tab:headline}
\begin{tabular}{@{}llllll@{}}
\toprule
Config & Accuracy & 95\,\% CI & Macro-F1 & MCC & Latency (ms) \\
\midrule
A (YOLO)       & \textbf{0.9615} & [0.9486, 0.9732] & 0.9618 & 0.9596 & 15.4  \\
B (YOLO+rules) & 0.9572          & [0.9443, 0.9700] & 0.9574 & 0.9551 & 444.4 \\
C (rules only) & 0.1338          & [0.1124, 0.1563] & 0.0771 & 0.0962 & 392.3 \\
\bottomrule
\end{tabular}
\end{table}

\noindent The A$\to$B gap is $-$0.43\,pp. SE\,$\approx$\,0.006; CIs overlap substantially.

\subsection{Expected Monetary Loss}

\begin{table}[h]
\centering
\caption{Expected monetary loss comparison.}
\label{tab:eml}
\begin{tabular}{@{}llll@{}}
\toprule
Config & Total EML (\rupee/ha) & Critical-disease EML (\rupee/ha) & $\Delta$ vs A \\
\midrule
A & 294.33    & 154.32    & ---      \\
B & 2{,}769.06 & 1{,}305.36 & +840.8\,\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Override Analysis}

The rule layer changes the CNN's top-1 prediction on \textbf{7 of 934 images}:
0 rescues, 4 corruptions, 3 neutral. Zero rescues from 7 overrides.
The rule engine fires on images where the CNN is already confident and correct.

\subsection{McNemar Test}

\begin{table}[h]
\centering
\caption{Paired McNemar test results.}
\label{tab:mcnemar}
\begin{tabular}{@{}lllll@{}}
\toprule
Pair  & $n_{\text{discordant}}$ & $\chi^2$ & $p$-value & Significant? \\
\midrule
A vs B & 4   & 2.25   & 0.134             & No  \\
A vs C & 777 & 767.03 & $<10^{-300}$      & Yes \\
B vs C & 773 & 763.03 & $<10^{-300}$      & Yes \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Per-Class Holm--Bonferroni Correction}

Per-class continuity-corrected McNemar tests with Holm--Bonferroni correction
across all 21 classes (Config~A vs~B, $n=933$) yield
\textbf{0/21 significant classes} ($\alpha=0.05$). The rule engine neither
reliably helps nor hurts any individual class. Full results are in
\texttt{evaluate/results/holm\_bonferroni\_perclass.csv}.

% ── 6. Threats to Validity ───────────────────────────────────
\section{Threats to Validity}

\begin{enumerate}
  \item \textbf{Small test set.} $n=934$ limits detectable effect size to
        $\geq\!\sim$0.6\,pp aggregate and $\geq\!\sim$0.06 F1 per class.
        A dataset of $n\in[2000,5000]$ would be needed to detect smaller effects.

  \item \textbf{Single backbone.} All results use YOLOv8n-cls (1.4M params).
        The finding may not transfer to weaker backbones where the CNN leaves
        more room for rules to add value. Multi-backbone comparison is planned
        (\texttt{configs/matrix/paper2.yaml}: 45 cells, 5 backbones $\times$
        3 geographies $\times$ 3 folds) but not yet executed.

  \item \textbf{Lab-quality imagery.} The evaluation set consists of
        smartphone leaf photographs, not field-deployed drone imagery.

  \item \textbf{Single seed.} Training used seed~42 only.

  \item \textbf{Cost parameters.} EML depends on the chosen cost matrix;
        different assumptions would change the 840\,\% headline. The cost
        table is committed at \texttt{configs/economics/india\_2025.yaml}.
\end{enumerate}

% ── 7. Conclusion ────────────────────────────────────────────
\section{Conclusion}

On 934 stratified smartphone images across 21 rice- and wheat-disease
classes, a hand-authored symbolic rule layer bolted onto a 96\,\%-accurate
YOLOv8n-cls backbone changes 4 predictions in the wrong direction, is not
significant under paired McNemar ($p=0.134$), imposes a 29$\times$ latency
tax, and inflates expected monetary loss by 840\,\% under a
smallholder-calibrated cost matrix. Per-class Holm--Bonferroni correction
confirms 0/21 classes show significant A-vs-B difference. At the accuracy
ceilings modern lightweight CNNs reach on curated leaf-photo datasets, the
neuro-symbolic hybrid pattern does not pay its rent.

\paragraph{Reproducibility.}
All prediction CSVs, bootstrap CI files, and evaluation scripts are committed
to the public repository (\url{\codeurl}). Model weights and dataset are
archived at \url{https://doi.org/\zenododoi}.

% ── References ───────────────────────────────────────────────
\bibliographystyle{splncs04}
\bibliography{references}

\end{document}
